---
title: "ML Creative Project 3"
subtitle: "Do Industry Classifiers Group Stocks Better Than K-Means?"
author: "Group 8 | University of South Florida"
date: today
format:
  revealjs:
    theme: [default]
    slide-number: true
    toc: true
    code-fold: true
    code-line-numbers: false
    chalkboard: true
    incremental: false
    transition: fade
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
editor: visual
---

```{r setup}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 4.8
)

set.seed(6776)

library(tidyverse)
library(lubridate)
library(DBI)
library(RPostgres)
library(dbplyr)
library(slider)
library(glmnet)
library(rsample)
library(yardstick)
library(broom)
library(mclust)
library(isotree)
library(knitr)
```

## Revised Question

**Old direction:** compare GICS sectors to Fama-French factor exposures.

**Issue:** factor exposures are risk explanations, not labels. If high exposure groups earn high returns, that may just reflect risk compensation.

**Revised direction:** compare **GICS**, **SIC**, and **K-means clusters** as stock grouping systems.

**Main question:** Which grouping system best captures return behavior and abnormal-return signals out of sample?

::: notes
This directly responds to the professor's feedback: do not treat Fama-French factors as the classifier benchmark. Use them as risk controls and lagged forecasting inputs only.
:::

## Professor Feedback Built Into the Design

1.  **No look-ahead in Fama-French data**\
    Use lagged Fama-French factors and lagged rolling exposures only.

2.  **Separate risk premium from anomaly**\
    Evaluate future returns as **risk-adjusted abnormal returns**, not just raw returns.

3.  **Better classifier comparison**\
    Compare GICS to SIC and unsupervised K-means clusters.

## Final Deliverable Format

This file is a **Quarto RevealJS presentation**.

Render with:

```{bash}
#| eval: false
quarto render ML_Creative_Project_3_Revised_Quarto_R.qmd
```

Optional report version:

```{bash}
#| eval: false
quarto render ML_Creative_Project_3_Revised_Quarto_R.qmd --to html
quarto render ML_Creative_Project_3_Revised_Quarto_R.qmd --to pdf
```

## Data Sources

| Data | Source | Main Use |
|------------------------|------------------------|------------------------|
| CRSP monthly stock returns | WRDS / CRSP | Return panel and delisting-adjusted returns |
| Company identifiers | WRDS / CRSP + Compustat | GICS and SIC classifications |
| Fama-French 5 factors | Ken French Data Library | Risk adjustment and lagged controls |
| Optional macro variables | FRED | Robustness only |

## Analysis Window

```{r config}
#| label: config
#| eval: false

start_date <- ymd("2000-01-01")
end_date   <- ymd("2024-12-31")

train_end  <- ymd("2017-12-31")
valid_end  <- ymd("2020-12-31")
test_start <- ymd("2021-01-01")

# Professor's data-hiatus concern:
# We only allow Fama-French information that would have been available with delay.
ff_lag_months <- 2
rolling_window_months <- 36
minimum_window_months <- 24
```

## WRDS Connection

```{r wrds-connection}
#| label: wrds-connection
#| eval: false

wrds_con <- dbConnect(
  RPostgres::Postgres(),
  host = "wrds-pgdata.wharton.upenn.edu",
  port = 9737,
  dbname = "wrds",
  sslmode = "require",
  user = Sys.getenv("WRDS_USER"),
  password = Sys.getenv("WRDS_PASSWORD")
)
```

## Pull CRSP Returns

```{r crsp-query}
#| label: crsp-query
#| eval: false

crsp_msf <- tbl(wrds_con, in_schema("crsp", "msf"))
crsp_names <- tbl(wrds_con, in_schema("crsp", "msenames"))

returns_raw <- crsp_msf |>
  filter(date >= !!start_date, date <= !!end_date) |>
  select(permno, date, ret, dlret, prc, shrout) |>
  collect() |>
  mutate(
    date = floor_date(as.Date(date), "month"),
    ret = as.numeric(ret),
    dlret = as.numeric(dlret),
    ret_adj = coalesce(ret, 0) + coalesce(dlret, 0) + coalesce(ret, 0) * coalesce(dlret, 0),
    mkt_cap = abs(as.numeric(prc)) * as.numeric(shrout)
  ) |>
  filter(!is.na(ret_adj), mkt_cap > 0)
```

## Pull SIC Codes

```{r sic-query}
#| label: sic-query
#| eval: false

sic_panel <- crsp_names |>
  select(permno, namedt, nameendt, siccd, exchcd, shrcd) |>
  collect() |>
  mutate(
    namedt = as.Date(namedt),
    nameendt = as.Date(nameendt),
    nameendt = coalesce(nameendt, ymd("2099-12-31")),
    sic2 = floor(as.numeric(siccd) / 100)
  )
```

## Pull GICS Codes

```{r gics-query}
#| label: gics-query
#| eval: false

# Depending on WRDS access, GICS may be available from Compustat company/security tables.
# First inspect available tables if needed:
# tbl(wrds_con, I("information_schema.tables")) |>
#   filter(table_schema %in% c("comp", "comp_na_daily_all")) |>
#   select(table_schema, table_name) |>
#   collect()

comp_company <- tbl(wrds_con, in_schema("comp", "company"))

company_gics <- comp_company |>
  select(gvkey, conm, sic, gsector, ggroup, gind, gsubind) |>
  collect() |>
  mutate(
    gics_sector = as.character(gsector),
    sic2_comp = floor(as.numeric(sic) / 100)
  )
```

## Fama-French Factors

```{r fama-french}
#| label: fama-french
#| eval: false

ff_url <- "https://mba.tuck.dartmouth.edu/pages/faculty/ken.french/ftp/F-F_Research_Data_5_Factors_2x3_CSV.zip"

ff_zip <- tempfile(fileext = ".zip")
download.file(ff_url, ff_zip, mode = "wb")
ff_file <- unzip(ff_zip, list = TRUE)$Name[1]
ff_text <- readLines(unz(ff_zip, ff_file))

start_row <- which(str_detect(ff_text, "^,Mkt-RF")) + 1
end_row <- which(str_detect(ff_text, "^ Annual Factors")) - 1

ff <- read_csv(
  paste(ff_text[start_row:end_row], collapse = "\n"),
  col_names = c("yyyymm", "mkt_rf", "smb", "hml", "rmw", "cma", "rf"),
  show_col_types = FALSE
) |>
  mutate(
    date = ymd(paste0(yyyymm, "01")),
    across(c(mkt_rf, smb, hml, rmw, cma, rf), ~ as.numeric(.x) / 100)
  ) |>
  select(date, mkt_rf, smb, hml, rmw, cma, rf) |>
  arrange(date)

# Data-hiatus design: factors are lagged by ff_lag_months.
ff_lagged <- ff |>
  arrange(date) |>
  mutate(
    across(
      c(mkt_rf, smb, hml, rmw, cma, rf),
      ~ lag(.x, ff_lag_months),
      .names = "{.col}_lag"
    )
  )
```

## Build Modeling Panel

```{r panel}
#| label: panel
#| eval: false

panel <- returns_raw |>
  left_join(ff_lagged, by = "date") |>
  mutate(
    ret_excess = ret_adj - rf_lag,
    forward_excess_1m = lead(ret_excess, 1),
    forward_excess_3m = slide_dbl(lead(ret_excess, 1), prod, .before = 0, .after = 2, .complete = TRUE) - 1,
    forward_excess_6m = slide_dbl(lead(ret_excess, 1), prod, .before = 0, .after = 5, .complete = TRUE) - 1
  ) |>
  group_by(permno) |>
  arrange(date, .by_group = TRUE) |>
  mutate(
    size_lag = lag(log(mkt_cap)),
    mom_12_2 = lag(slide_dbl(1 + ret_adj, prod, .before = 11, .after = -2, .complete = TRUE) - 1),
    vol_12 = lag(slide_dbl(ret_adj, sd, .before = 11, .complete = TRUE))
  ) |>
  ungroup()
```

## Rolling Factor Exposures

```{r rolling-exposures}
#| label: rolling-exposures
#| eval: false

estimate_exposure <- function(df) {
  df <- drop_na(df, ret_excess, mkt_rf_lag, smb_lag, hml_lag, rmw_lag, cma_lag)
  if (nrow(df) < minimum_window_months) return(tibble(alpha = NA_real_, beta_mkt = NA_real_, beta_smb = NA_real_, beta_hml = NA_real_, beta_rmw = NA_real_, beta_cma = NA_real_))

  fit <- lm(ret_excess ~ mkt_rf_lag + smb_lag + hml_lag + rmw_lag + cma_lag, data = df)
  co <- coef(fit)

  tibble(
    alpha = unname(co["(Intercept)"]),
    beta_mkt = unname(co["mkt_rf_lag"]),
    beta_smb = unname(co["smb_lag"]),
    beta_hml = unname(co["hml_lag"]),
    beta_rmw = unname(co["rmw_lag"]),
    beta_cma = unname(co["cma_lag"])
  )
}

exposures <- panel |>
  arrange(permno, date) |>
  group_by(permno) |>
  group_modify(~ {
    .x |>
      mutate(row_id = row_number()) |>
      group_by(row_id) |>
      group_modify(~ {
        this_date <- .x$date[1]
        hist <- panel |>
          filter(permno == .x$permno[1], date < this_date, date >= this_date %m-% months(rolling_window_months))
        estimate_exposure(hist)
      }) |>
      ungroup() |>
      bind_cols(.x |> select(date))
  }) |>
  ungroup()

panel_features <- panel |>
  left_join(exposures, by = c("permno", "date")) |>
  drop_na(size_lag, mom_12_2, vol_12, beta_mkt, beta_smb, beta_hml, beta_rmw, beta_cma)
```

## Split Data

```{r split}
#| label: split
#| eval: false

train <- panel_features |> filter(date <= train_end)
valid <- panel_features |> filter(date > train_end, date <= valid_end)
test  <- panel_features |> filter(date >= test_start)
```

## K-Means Groups

```{r kmeans}
#| label: kmeans
#| eval: false

cluster_vars <- c("size_lag", "mom_12_2", "vol_12", "beta_mkt", "beta_smb", "beta_hml", "beta_rmw", "beta_cma")

scale_fit <- train |>
  select(all_of(cluster_vars)) |>
  scale()

k_grid <- 4:12
wss <- map_dbl(k_grid, ~ kmeans(scale_fit, centers = .x, nstart = 25)$tot.withinss)

chosen_k <- 8
km_fit <- kmeans(scale_fit, centers = chosen_k, nstart = 50)

train_clusters <- train |>
  mutate(cluster = factor(km_fit$cluster))

assign_cluster <- function(new_data, centers, vars) {
  x <- scale(new_data |> select(all_of(vars)),
             center = attr(scale_fit, "scaled:center"),
             scale = attr(scale_fit, "scaled:scale"))
  distances <- as.matrix(dist(rbind(centers, x)))[-(1:nrow(centers)), 1:nrow(centers), drop = FALSE]
  factor(max.col(-distances))
}

valid_clusters <- valid |> mutate(cluster = assign_cluster(valid, km_fit$centers, cluster_vars))
test_clusters  <- test  |> mutate(cluster = assign_cluster(test,  km_fit$centers, cluster_vars))
```

## Classifier Comparison

```{r classifier-comparison}
#| label: classifier-comparison
#| eval: false

comparison_data <- test_clusters |>
  filter(!is.na(gics_sector), !is.na(sic2), !is.na(cluster))

ari_gics_sic <- adjustedRandIndex(comparison_data$gics_sector, comparison_data$sic2)
ari_gics_km  <- adjustedRandIndex(comparison_data$gics_sector, comparison_data$cluster)
ari_sic_km   <- adjustedRandIndex(comparison_data$sic2, comparison_data$cluster)

classifier_overlap <- tibble(
  comparison = c("GICS vs SIC", "GICS vs K-means", "SIC vs K-means"),
  adjusted_rand_index = c(ari_gics_sic, ari_gics_km, ari_sic_km)
)

classifier_overlap |> kable(digits = 3)
```

## Heatmap: GICS vs K-Means

```{r heatmap}
#| label: heatmap
#| eval: false

comparison_data |>
  count(gics_sector, cluster) |>
  group_by(gics_sector) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(cluster, gics_sector, fill = pct)) +
  geom_tile() +
  scale_fill_continuous(labels = scales::percent) +
  labs(
    title = "How K-means clusters map to GICS sectors",
    x = "K-means cluster",
    y = "GICS sector",
    fill = "Within-sector share"
  ) +
  theme_minimal()
```

## Prediction Design

Prediction target: **next-month excess return**.

Models tested:

1.  Controls only\
2.  Controls + GICS\
3.  Controls + SIC\
4.  Controls + K-means cluster\
5.  Controls + GICS + SIC + K-means

## Elastic Net Utility

```{r elastic-net-utility}
#| label: elastic-net-utility
#| eval: false

fit_glmnet_model <- function(train_df, test_df, group_var = NULL) {
  base_vars <- c("size_lag", "mom_12_2", "vol_12", "beta_mkt", "beta_smb", "beta_hml", "beta_rmw", "beta_cma")
  rhs <- if (is.null(group_var)) {
    paste(base_vars, collapse = " + ")
  } else {
    paste(c(base_vars, group_var), collapse = " + ")
  }

  f <- as.formula(paste("forward_excess_1m ~", rhs))

  train_clean <- train_df |> drop_na(forward_excess_1m, all_of(base_vars))
  test_clean  <- test_df  |> drop_na(forward_excess_1m, all_of(base_vars))

  x_train <- model.matrix(f, train_clean)[, -1]
  y_train <- train_clean$forward_excess_1m
  x_test <- model.matrix(f, test_clean)[, -1]

  fit <- cv.glmnet(x_train, y_train, alpha = 0.5, nfolds = 5)
  pred <- as.numeric(predict(fit, newx = x_test, s = "lambda.min"))

  tibble(
    actual = test_clean$forward_excess_1m,
    pred = pred,
    direction_actual = actual > 0,
    direction_pred = pred > 0
  )
}
```

## Out-of-Sample Accuracy

```{r prediction-results}
#| label: prediction-results
#| eval: false

train_model <- train_clusters
valid_model <- valid_clusters
test_model  <- test_clusters

predictions <- list(
  Controls = fit_glmnet_model(train_model, test_model),
  GICS = fit_glmnet_model(train_model, test_model, "gics_sector"),
  SIC = fit_glmnet_model(train_model, test_model, "sic2"),
  KMeans = fit_glmnet_model(train_model, test_model, "cluster")
)

model_metrics <- imap_dfr(predictions, ~ {
  tibble(
    model = .y,
    rmse = sqrt(mean((.x$actual - .x$pred)^2, na.rm = TRUE)),
    mae = mean(abs(.x$actual - .x$pred), na.rm = TRUE),
    r2_oos = 1 - sum((.x$actual - .x$pred)^2, na.rm = TRUE) / sum((.x$actual - mean(.x$actual, na.rm = TRUE))^2, na.rm = TRUE),
    directional_accuracy = mean(.x$direction_actual == .x$direction_pred, na.rm = TRUE)
  )
})

model_metrics |> arrange(desc(r2_oos)) |> kable(digits = 4)
```

## Why Raw Return Is Not Enough

Finding that a group earns higher raw returns does **not** prove anomaly.

A group may simply have higher exposure to compensated risk factors.

Therefore, anomaly testing focuses on:

$$\alpha_{i,t+h} = R_{i,t+h} - R_{f,t+h} - \hat{\beta}_{i,t}'F_{t+h}$$

## Risk-Adjusted Forward Alpha

```{r risk-adjusted-alpha}
#| label: risk-adjusted-alpha
#| eval: false

test_alpha <- test_clusters |>
  left_join(ff, by = "date", suffix = c("", "_realized")) |>
  mutate(
    expected_factor_return = beta_mkt * mkt_rf + beta_smb * smb + beta_hml * hml + beta_rmw * rmw + beta_cma * cma,
    abnormal_1m = forward_excess_1m - expected_factor_return
  )
```

## Isolation Forest Within Each Classifier

```{r isolation-forest}
#| label: isolation-forest
#| eval: false

score_isolation_by_group <- function(df, group_var, vars = cluster_vars) {
  df |>
    group_by(across(all_of(group_var))) |>
    group_modify(~ {
      if (nrow(.x) < 50) {
        return(.x |> mutate(anomaly_score = NA_real_, anomaly_flag = FALSE))
      }
      x <- .x |> select(all_of(vars)) |> drop_na()
      if (nrow(x) < 50) {
        return(.x |> mutate(anomaly_score = NA_real_, anomaly_flag = FALSE))
      }
      fit <- isolation.forest(as.matrix(x), ntrees = 100, sample_size = min(256, nrow(x)))
      scores <- predict(fit, as.matrix(x), type = "score")
      threshold <- quantile(scores, 0.95, na.rm = TRUE)
      .x |> mutate(anomaly_score = scores, anomaly_flag = scores >= threshold)
    }) |>
    ungroup()
}

anom_gics <- score_isolation_by_group(test_alpha, "gics_sector") |> mutate(classifier = "GICS")
anom_sic  <- score_isolation_by_group(test_alpha, "sic2") |> mutate(classifier = "SIC")
anom_km   <- score_isolation_by_group(test_alpha, "cluster") |> mutate(classifier = "KMeans")

anomalies <- bind_rows(anom_gics, anom_sic, anom_km)
```

## Anomaly Test

```{r anomaly-test}
#| label: anomaly-test
#| eval: false

anomaly_summary <- anomalies |>
  group_by(classifier, anomaly_flag) |>
  summarize(
    n = n(),
    mean_alpha_1m = mean(abnormal_1m, na.rm = TRUE),
    median_alpha_1m = median(abnormal_1m, na.rm = TRUE),
    hit_rate = mean(abnormal_1m > 0, na.rm = TRUE),
    .groups = "drop"
  )

anomaly_summary |> kable(digits = 4)
```

## Bootstrap Confidence Intervals

```{r bootstrap-ci}
#| label: bootstrap-ci
#| eval: false

bootstrap_ci <- function(x, statistic = mean, n_boot = 1000) {
  x <- x[is.finite(x)]
  sims <- replicate(n_boot, statistic(sample(x, length(x), replace = TRUE)))
  quantile(sims, c(0.025, 0.975), na.rm = TRUE)
}

ci_table <- anomalies |>
  filter(anomaly_flag) |>
  group_by(classifier) |>
  summarize(
    mean_alpha = mean(abnormal_1m, na.rm = TRUE),
    ci_low = bootstrap_ci(abnormal_1m)[1],
    ci_high = bootstrap_ci(abnormal_1m)[2],
    .groups = "drop"
  )

ci_table |> kable(digits = 4)
```

## Final Evaluation Questions

1.  How much do GICS, SIC, and K-means overlap?
2.  Which grouping system gives better out-of-sample return prediction?
3.  Which grouping system produces anomaly flags with stronger future alpha?
4.  Are anomalies classifier-specific, or do all systems flag the same firms?

## Expected Interpretation

If K-means beats GICS and SIC, the result suggests that data-driven groupings capture economically meaningful similarities not fully reflected in standard industry codes.

If GICS or SIC wins, the result supports the idea that traditional industry classifications already summarize most relevant similarities.

If no classifier improves alpha prediction, the honest conclusion is that group labels do not add much beyond lagged firm characteristics and factor exposures.

## Project Guardrails

- Use only lagged factors and historical exposures.
- Train clusters only on the training set.
- Assign validation/test firms to fixed training clusters.
- Report raw return results and risk-adjusted alpha results separately.
- Treat anomaly results as exploratory unless confidence intervals are clearly nonzero.

## Backup Slide: What Changed From the Blueprint?

| Original Plan | Revised Plan |
|------------------------------------|------------------------------------|
| Python | R + Quarto |
| Compare GICS to FF exposures | Compare GICS to SIC and K-means |
| Raw return anomaly framing | Risk-adjusted alpha framing |
| FF factors available contemporaneously | FF factors lagged by `ff_lag_months` |
| Isolation forest only as unusual stock detector | Isolation forest evaluated by future abnormal returns |

## Backup Slide: One-Sentence Pitch

We test whether machine-learned stock clusters classify firms more usefully than traditional GICS or SIC industry labels, using strictly lagged factor information and evaluating success by out-of-sample prediction and future risk-adjusted alpha.