---
title: "ML Creative Project 3 — Three-Way Equity Classifier Comparison"
subtitle: "Phase 1 + Phase 2 — Data Assembly and Feature Engineering"
author: "Group 8 | University of South Florida"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
    embed-resources: true
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
execute:
  warning: false
  message: false
  cache: false
---

```{r setup}
#| include: true

# ============================================================================
# PROJECT-WIDE CONFIG
# Set dev_mode = TRUE for fast iteration on a 50-stock subset.
# Set dev_mode = FALSE for the full S&P 500 universe.
# Everything else flows from this single switch.
# ============================================================================

CFG <- list(
  dev_mode      = TRUE,                         # <<< toggle here
  dev_n_stocks  = 50,
  start_date    = as.Date("2014-01-01"),
  end_date      = as.Date("2024-12-31"),
  train_start   = as.Date("2018-01-01"),
  train_end     = as.Date("2022-12-31"),
  test_start    = as.Date("2023-01-01"),
  test_end      = as.Date("2024-12-31"),
  loading_window_months = 36,
  loading_min_obs       = 24,
  loading_lag_months    = 2,
  macro_lag_months      = 1,
  seed = 8675309
)
assign("CFG", CFG, envir = .GlobalEnv)

source("R/utils.R")
source("R/01_pull_crsp.R")
source("R/02_pull_french.R")
source("R/03_pull_fred.R")
source("R/04_build_panel.R")
source("R/05_features.R")

library(ggplot2)
library(knitr)
library(scales)

theme_set(theme_minimal(base_size = 11))
```

# Overview

This document executes Phases 1 and 2 of the project: data assembly and feature engineering. Subsequent phases (clustering, comparison battery, anomaly detection) read from the parquet files produced here.

**Mode:** `r if (CFG$dev_mode) "DEVELOPMENT — small subset for fast iteration" else "FULL — entire S&P 500 universe"`

**Universe:** Historical S&P 500 constituents, point-in-time, including delisted names.
**Date range:** `r CFG$start_date` to `r CFG$end_date`.

# Phase 1 — Data Assembly

## Pull CRSP

Historical S&P 500 membership, monthly returns adjusted for delistings, stock header info (ticker, SIC), and GICS sector via the CCM linkage.

```{r pull-crsp}
crsp <- pull_crsp_all()
```

**S&P 500 membership spans:**

```{r}
crsp$members |>
  summarise(
    spans = n(),
    permnos = n_distinct(permno),
    earliest_start = min(start),
    latest_end = max(end_date)
  ) |>
  kable()
```

**Monthly return panel:**

```{r}
crsp$monthly |>
  summarise(
    rows = n(),
    permnos = n_distinct(permno),
    months = n_distinct(date_eom),
    delisting_obs = sum(!is.na(dlret)),
    pct_ret_present = mean(!is.na(ret))
  ) |>
  kable(digits = 3)
```

## Pull Ken French

Five factors plus risk-free rate (monthly), and the FF12 industry definitions (SIC ranges).

```{r pull-french}
french <- pull_french_all()
```

**FF5 factors — head and tail:**

```{r}
bind_rows(head(french$ff5, 3), tail(french$ff5, 3)) |>
  kable(digits = 4)
```

**FF12 industry definitions:**

```{r}
french$ff12_def |>
  group_by(industry_num, industry_code, industry_name) |>
  summarise(n_sic_ranges = n(), .groups = "drop") |>
  arrange(industry_num) |>
  kable()
```

## Pull FRED

10-year Treasury yield, Fed funds rate, VIX (daily, aggregated to end-of-month).

```{r pull-fred}
fred <- pull_fred_all()
```

```{r}
fred |>
  summarise(
    months = n(),
    earliest = min(date_eom),
    latest = max(date_eom),
    pct_ten_y = mean(!is.na(ten_y)),
    pct_ffr = mean(!is.na(ffr)),
    pct_vix = mean(!is.na(vix))
  ) |>
  kable(digits = 3)
```

## Build the panel

Merge everything. The panel is one row per (permno, end-of-month) for every month in which the stock was an S&P 500 constituent.

```{r build-panel}
panel <- build_panel(crsp, french, fred)
```

**Panel summary:**

```{r}
panel |>
  summarise(
    rows = n(),
    permnos = n_distinct(permno),
    months = n_distinct(date_eom),
    pct_ret_excess = mean(!is.na(ret_excess)),
    pct_gics = mean(!is.na(gsector)),
    pct_ff12 = mean(!is.na(ff12_num)),
    pct_macro = mean(!is.na(ten_y) & !is.na(ffr) & !is.na(vix))
  ) |>
  kable(digits = 3)
```

**GICS sector distribution (most recent year):**

```{r}
recent <- max(panel$date_eom)
panel |>
  filter(date_eom == recent, !is.na(gics_sector_name)) |>
  count(gics_sector_name, sort = TRUE) |>
  kable()
```

**FF12 industry distribution (most recent year):**

```{r}
panel |>
  filter(date_eom == recent, !is.na(ff12_code)) |>
  count(ff12_num, ff12_code, sort = TRUE) |>
  kable()
```

# Phase 2 — Feature Engineering

The features panel is the output of strict, audited transformations on the raw panel:

1. **Rolling FF5 loadings** estimated on a 36-month window ending at $t-2$.
2. **Realized alpha** at $t$ = $r^e_t - \sum_k \beta_{i,k} F_{k,t}$, where $\beta$ comes from the $t-2$ window (label only, never a feature).
3. **Cross-sectional z-scores** of loadings — at each date, z-score across the cross-section.
4. **Standard controls**: 12-1 momentum, 12m volatility, log market cap (lagged 1m).
5. **Macro lags**: 10Y, FFR, VIX all lagged 1m.
6. **Forward target**: $\alpha_{i,t+1}$ obtained by leading realized alpha by 1 month within each stock.

```{r features}
feats <- build_features(panel)
```

## Coverage diagnostics

```{r}
feats |>
  summarise(
    rows = n(),
    pct_loadings = mean(!is.na(beta_mkt)),
    pct_realized_alpha = mean(!is.na(alpha_realized)),
    pct_fwd_target = mean(!is.na(alpha_fwd_1m)),
    modelable = mean(!is.na(alpha_fwd_1m) & !is.na(beta_mkt) & !is.na(mom_12_1))
  ) |>
  kable(digits = 3)
```

## Anti-leakage tests

```{r leak-tests}
leak_results <- run_leakage_tests(feats)
```

The critical test: `alpha_fwd_1m` at $(i, t)$ must equal `alpha_realized` at $(i, t+1)$ for every paired row. This is reported above (`TRUE` = pass).

## Distributions of key variables

```{r}
#| fig-height: 3
#| fig-width: 8
feats |>
  filter(!is.na(alpha_realized)) |>
  ggplot(aes(alpha_realized)) +
  geom_histogram(bins = 60, fill = "#2E75B6", alpha = 0.85) +
  labs(title = "Realized alpha (monthly)",
       x = "alpha = excess return - β·FF5",
       y = "count")
```

```{r}
#| fig-height: 4
#| fig-width: 8
feats |>
  filter(!is.na(beta_mkt)) |>
  select(beta_mkt, beta_smb, beta_hml, beta_rmw, beta_cma) |>
  pivot_longer(everything(), names_to = "factor", values_to = "loading") |>
  ggplot(aes(loading)) +
  geom_histogram(bins = 50, fill = "#1F4E79", alpha = 0.85) +
  facet_wrap(~ factor, scales = "free", ncol = 5) +
  labs(title = "FF5 factor loadings (rolling 36m, ending t-2)",
       x = NULL, y = NULL)
```

## Sample of feature rows

```{r}
feats |>
  filter(!is.na(alpha_fwd_1m), !is.na(beta_mkt)) |>
  slice_sample(n = 8) |>
  select(permno, date_eom, ticker, gics_sector_name, ff12_code,
         alpha_realized, alpha_fwd_1m,
         beta_mkt, beta_hml,
         z_beta_mkt, z_beta_hml,
         mom_12_1, vol_12m, log_mktcap_lag, ten_y_lag) |>
  kable(digits = 4)
```

# Outputs

Phase 1 + 2 wrote the following files:

```{r}
list.files(c(PATHS$raw, PATHS$processed),
           pattern = "\\.parquet$", full.names = TRUE) |>
  tibble(file = _) |>
  mutate(size_kb = round(file.size(file) / 1024, 1)) |>
  kable()
```

Phase 3 (build the three classifiers) reads from `data/processed/features*.parquet`.
