---
title: "Three-Way Equity Classifier Comparison"
subtitle: "GICS Sectors vs. SIC/FF12 Industries vs. K-Means Co-Movement Clusters"
author: "Group 8 | University of South Florida"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
    embed-resources: true
    fig-width: 8
    fig-height: 4.5
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
execute:
  warning: false
  message: false
  cache: false
---

```{r setup}
#| include: false

# ============================================================================
# Project-wide config. Toggle dev_mode for fast iteration on a 50-stock
# subset. ff_lag_months addresses the FF data publication delay so the
# pipeline could be deployed in real time.
# ============================================================================

CFG <- list(
  dev_mode      = TRUE,
  dev_n_stocks  = 50,
  start_date    = as.Date("2014-01-01"),
  end_date      = as.Date("2024-12-31"),
  train_start   = as.Date("2018-01-01"),
  train_end     = as.Date("2022-12-31"),
  test_start    = as.Date("2023-01-01"),
  test_end      = as.Date("2024-12-31"),
  loading_window_months = 36,
  loading_min_obs       = 24,
  loading_lag_months    = 2,    # rolling-beta window ends at t-2
  ff_lag_months         = 2,    # FF5 publication lag
  macro_lag_months      = 1,    # FRED macro lag
  seed = 8675309
)
assign("CFG", CFG, envir = .GlobalEnv)

source("R/utils.R")
source("R/01_pull_crsp.R")
source("R/02_pull_french.R")
source("R/03_pull_fred.R")
source("R/04_build_panel.R")
source("R/05_features.R")
source("R/06_clustering.R")
source("R/07_compare.R")

library(ggplot2)
library(knitr)
library(scales)

local({
  qmd_dir <- tryCatch(
    dirname(normalizePath(knitr::current_input(dir = TRUE), winslash = "/")),
    error = function(e) getwd()
  )
  candidates <- unique(c(
    file.path(qmd_dir, "credentials.R"),
    file.path(getwd(), "credentials.R"),
    "credentials.R"
  ))
  hit <- candidates[file.exists(candidates)][1]
  if (is.na(hit)) {
    stop("credentials.R not found. Looked in:\n  ",
         paste(candidates, collapse = "\n  "),
         "\nCreate credentials.R with `wrds_user <- \"...\"` and ",
         "`wrds_password <- \"...\"` next to report.qmd.")
  }
  sys.source(hit, envir = .GlobalEnv)
})

theme_set(theme_minimal(base_size = 11))
```

# Executive summary

This project asks whether data-driven groupings of S&P 500 stocks classify
firms more coherently than the GICS industry standard, and whether one
grouping scheme produces better return-anomaly signals than another. We
compare three classification systems head-to-head:

1. **GICS sectors** — the eleven Global Industry Classification Standard
   sectors that the index providers use.
2. **SIC-based FF12 industries** — Ken French's twelve-industry partition,
   which is built directly from SIC codes and is the academic-finance
   counterpart to GICS.
3. **K-Means clusters on return co-movement** — an unsupervised partition
   that groups stocks whose monthly excess returns move together over a
   rolling sixty-month window, refit annually with label matching.

All three schemes are fed through the same downstream pipeline on the same
data, and we compare them on four dimensions: composition overlap,
within-group homogeneity, predictive accuracy on one-month-forward alpha,
and anomaly-detection behaviour.

Two design choices, made in response to feedback on an earlier draft of
this project, deserve to be flagged at the top:

- **The target is alpha, not raw return.** Predicting realised return from
  factor exposures would mostly recover the standard risk premia
  (high-beta stocks earn higher returns on average) — a tautology, not a
  finding. The economically interesting quantity is the residual after
  factor exposure is priced out, i.e. realised alpha. We predict
  $\alpha_{i,t+1}$ throughout.
- **FF factors enter as features only with a publication lag.** Ken
  French's data library publishes FF5 with a multi-month delay, so a
  forecast that consumes contemporaneous FF values would be unusable in
  real time. Every feature derived from FF returns is lagged by
  `r CFG$ff_lag_months` months. (Realised alpha at month $t$ is computed
  from $F_t$, but it is a *label* — never a feature input — so this is
  internally consistent.)

The dataset and pipeline are built with the dev-mode toggle in `CFG`. The
HTML version of this report renders to the full S&P 500 universe when
`CFG$dev_mode = FALSE`; we leave dev mode on here for reproducibility on
modest hardware. Numbers and figures will tighten when run on the full
universe.

# 1. Research question

Do data-driven groupings of S&P 500 stocks classify firms more coherently
than GICS sectors, and do they better surface forward-alpha anomalies?

This restates the question we posed in the original project blueprint in
two ways. First, we no longer compare an industry classification (GICS) to
a *risk* taxonomy (FF5 loadings), because those are incommensurable. The
right comparison set for an industry classification is other industry
classifications — GICS, SIC/FF12, and an unsupervised industry-style
clustering. Second, we judge "better classification" not just on
predictive R² but also on within-group homogeneity (do members really
behave alike?) and on anomaly-flagging quality (do flagged outliers
subsequently earn different forward alpha?).

# 2. Methods

## 2.1 The three classification schemes

**GICS sectors.** Eleven sectors assigned per stock-month via the
CRSP/Compustat Merged (CCM) linkage. Membership is point-in-time: a stock
that switches sectors mid-sample carries the new sector going forward.

**SIC/FF12.** Each stock-month's SIC code (from CRSP `stocknames`,
respecting historical name ranges) is mapped to one of twelve Ken French
industries. SIC codes that fall outside any range get assigned to "Other."

**K-Means clusters on return co-movement.** At the end of each calendar
year, we take the prior sixty months of monthly excess returns. For each
stock with at least thirty-six valid months we form a feature vector
equal to that stock's column of the cross-sectional correlation matrix
— i.e. each stock is described by its correlation with every other stock
in the panel. We reduce to ten principal components and fit K-Means for
$k \in \{2, \ldots, 10\}$, picking $k$ that maximises the mean
silhouette. The fitted assignment is used for all twelve months of the
following calendar year. Cluster identities are kept stable across refits
by Hungarian-algorithm matching against the previous fit's labels.

## 2.2 Target: realised alpha

For each stock-month we compute realised alpha as the residual from the
five-factor model:
$$
\alpha_{i,t} \;=\; r^e_{i,t} \;-\; \sum_{k \in \{\text{Mkt},\text{SMB},\text{HML},\text{RMW},\text{CMA}\}} \hat\beta_{i,k}\, F_{k,t},
$$
where $r^e_{i,t}$ is excess return and $\hat\beta_{i,k}$ is estimated by
OLS on the 36-month window ending at month $t-2$. The forward target used
throughout is $\alpha_{i,t+1}$ — realised alpha one month ahead — attached
to the row at month $t$.

## 2.3 Features

The feature vector at month $t$ contains:

- Cross-sectionally z-scored FF5 betas (`z_beta_*`), ranking each stock
  against the cross-section at the same date.
- 12-1 momentum (cumulative return from $t-12$ to $t-2$), 12-month vol of
  excess returns, and log market cap lagged one month.
- Lagged macro: 10-year Treasury yield, Fed funds rate, VIX — each lagged
  one month.
- Lagged FF5 returns (`mkt_rf_lag`, `smb_lag`, `hml_lag`, `rmw_lag`,
  `cma_lag`) — each lagged `r CFG$ff_lag_months` months to honour the
  publication delay.

## 2.4 Evaluation

Train: January 2018 – December 2022 (after the 36-month rolling-beta
warm-up consumes 2014–2017). Holdout: January 2023 – December 2024. All
metrics report on the holdout.

# 3. Data assembly

## 3.1 Pull CRSP

Historical S&P 500 membership, monthly returns adjusted for delisting,
historical SIC codes, and GICS via the CCM linkage.

```{r pull-crsp}
crsp <- pull_crsp_all()
```

```{r}
crsp$members |>
  summarise(
    spans = n(),
    permnos = n_distinct(permno),
    earliest_start = min(start),
    latest_end = max(end_date)
  ) |>
  kable(caption = "S&P 500 membership spans")
```

```{r}
crsp$monthly |>
  summarise(
    rows = n(),
    permnos = n_distinct(permno),
    months = n_distinct(date_eom),
    delisting_obs = sum(!is.na(dlret)),
    pct_ret_present = mean(!is.na(ret))
  ) |>
  kable(digits = 3, caption = "Monthly return panel")
```

## 3.2 Pull Ken French

```{r pull-french}
french <- pull_french_all()
```

```{r}
bind_rows(head(french$ff5, 3), tail(french$ff5, 3)) |>
  kable(digits = 4, caption = "FF5 factors — head and tail")
```

```{r}
french$ff12_def |>
  group_by(industry_num, industry_code, industry_name) |>
  summarise(n_sic_ranges = n(), .groups = "drop") |>
  arrange(industry_num) |>
  kable(caption = "FF12 industry definitions")
```

## 3.3 Pull FRED

```{r pull-fred}
fred <- pull_fred_all()
```

```{r}
fred |>
  summarise(
    months = n(),
    earliest = min(date_eom),
    latest   = max(date_eom),
    pct_ten_y = mean(!is.na(ten_y)),
    pct_ffr   = mean(!is.na(ffr)),
    pct_vix   = mean(!is.na(vix))
  ) |>
  kable(digits = 3, caption = "FRED macro coverage")
```

## 3.4 Build the panel

```{r build-panel}
panel <- build_panel(crsp, french, fred)
```

```{r}
panel |>
  summarise(
    rows = n(),
    permnos = n_distinct(permno),
    months  = n_distinct(date_eom),
    pct_ret_excess = mean(!is.na(ret_excess)),
    pct_gics       = mean(!is.na(gsector)),
    pct_ff12       = mean(!is.na(ff12_num)),
    pct_macro      = mean(!is.na(ten_y) & !is.na(ffr) & !is.na(vix))
  ) |>
  kable(digits = 3, caption = "Master panel coverage")
```

```{r}
recent <- max(panel$date_eom)
panel |>
  filter(date_eom == recent, !is.na(gics_sector_name)) |>
  dplyr::count(gics_sector_name, sort = TRUE) |>
  kable(caption = paste0("GICS sector distribution at ", recent))
```

```{r}
panel |>
  filter(date_eom == recent, !is.na(ff12_code)) |>
  dplyr::count(ff12_num, ff12_code, sort = TRUE) |>
  kable(caption = paste0("FF12 industry distribution at ", recent))
```

# 4. Feature engineering

```{r features}
feats <- build_features(panel)
```

## 4.1 Coverage

```{r}
feats |>
  summarise(
    rows = n(),
    pct_loadings       = mean(!is.na(beta_mkt)),
    pct_realized_alpha = mean(!is.na(alpha_realized)),
    pct_fwd_target     = mean(!is.na(alpha_fwd_1m)),
    modelable          = mean(!is.na(alpha_fwd_1m) & !is.na(beta_mkt) &
                              !is.na(mom_12_1) & !is.na(mkt_rf_lag))
  ) |>
  kable(digits = 3, caption = "Feature panel coverage")
```

## 4.2 Anti-leakage tests

The critical invariant: `alpha_fwd_1m` at $(i, t)$ must equal
`alpha_realized` at $(i, t+1)$ for every paired row. We also verify that
the lagged FF5 columns match a manual lag operation.

```{r leak-tests}
leak_results <- run_leakage_tests(feats)
```

## 4.3 Distributions

```{r alpha-hist}
#| fig-cap: "Realised monthly alpha — symmetric around zero, fat-tailed."
feats |>
  filter(!is.na(alpha_realized)) |>
  ggplot(aes(alpha_realized)) +
  geom_histogram(bins = 60, fill = "#2E75B6", alpha = 0.85) +
  labs(x = expression(alpha == r^e - beta %.% F), y = "count")
```

```{r beta-hist}
#| fig-cap: "FF5 betas across the panel — mostly market-driven."
feats |>
  filter(!is.na(beta_mkt)) |>
  select(beta_mkt, beta_smb, beta_hml, beta_rmw, beta_cma) |>
  pivot_longer(everything(), names_to = "factor", values_to = "loading") |>
  ggplot(aes(loading)) +
  geom_histogram(bins = 50, fill = "#1F4E79", alpha = 0.85) +
  facet_wrap(~ factor, scales = "free", ncol = 5) +
  labs(x = NULL, y = NULL)
```

# 5. Clustering on return co-movement

K-Means is fit at the end of each year on the prior sixty months; the
assignment is applied to all twelve months of the following calendar year.

```{r clustering}
clust <- build_clusters_annual(panel)
```

```{r}
clust$diag |>
  kable(digits = 3, caption = "Cluster diagnostics by refit date")
```

```{r cluster-k-plot}
#| fig-cap: "Chosen k by refit year. Stability across refits is encouraging — drift in $k$ would suggest the structure is unstable."
clust$diag |>
  ggplot(aes(refit_date, k)) +
  geom_point(size = 2.5, colour = "#C00000") +
  geom_line(colour = "#C00000") +
  scale_y_continuous(breaks = 2:10) +
  labs(x = "refit date (end of year)", y = "chosen k")
```

# 6. Three-way comparison

```{r merge-clusters}
feats_g <- attach_clusters(feats, clust$assignments)
```

## 6.1 Composition

How much do the three groupings overlap?

```{r composition}
comp <- composition_compare(feats_g)
comp$ari |>
  kable(digits = 3, caption = "Adjusted Rand index between scheme pairs")
```

The adjusted Rand index measures partition similarity, with 1 = identical
and 0 = random. Values close to zero between GICS or FF12 and the K-Means
clusters mean the unsupervised partition is genuinely different from the
industry partitions, which is what we'd want — otherwise it adds nothing.

```{r contingency-gics-cluster}
comp$contingency$gics_vs_cluster |>
  as.data.frame.matrix() |>
  kable(caption = "GICS sector × K-Means cluster (most recent month)")
```

```{r contingency-ff12-cluster}
comp$contingency$ff12_vs_cluster |>
  as.data.frame.matrix() |>
  kable(caption = "FF12 industry × K-Means cluster (most recent month)")
```

## 6.2 Within-group homogeneity

A "good" classification produces internally similar members. We measure
this two ways: the cross-sectional standard deviation of realised alpha
within each group at each date (lower = more homogeneous in unexplained
returns), and the mean pairwise correlation of excess returns among
group members (higher = more co-moving).

```{r homogeneity}
homo <- homogeneity_compare(feats_g, panel)
homo |> kable(digits = 4, caption = "Within-group homogeneity by scheme")
```

If the K-Means scheme is doing its job, it should win on
`mean_within_corr` — that's what it was optimised for. The question is
whether GICS or FF12 closes that gap.

## 6.3 Predictive accuracy

Per-group Elastic Net regressions predict $\alpha_{i,t+1}$ from the
feature set. We report out-of-sample R², MAE, and directional accuracy on
the 2023–2024 holdout, weighted by group size.

```{r prediction}
pred <- prediction_compare(feats_g)
pred$aggregate |>
  arrange(desc(r2_oos_weighted)) |>
  kable(digits = 4, caption = "Aggregate prediction metrics by scheme")
```

```{r prediction-per-group, fig.height=6}
pred$per_group |>
  ggplot(aes(reorder(group, r2_oos), r2_oos, fill = scheme)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ scheme, scales = "free_y") +
  labs(x = NULL, y = "out-of-sample R²", fill = NULL,
       title = "OOS R² per group, by scheme") +
  scale_fill_manual(values = c(GICS = "#1F4E79", FF12 = "#2E75B6", Cluster = "#C00000")) +
  theme(legend.position = "none")
```

The pooled baseline (no grouping) gives a single all-stocks Elastic Net.
If any of the three schemes can beat the pooled model on R² *and*
directional accuracy, that scheme is adding information beyond what the
features carry on their own.

## 6.4 Anomaly detection

Isolation Forest is trained per group on the train window and scored on
the holdout; the most anomalous five percent are flagged. We then look at
average forward alpha at one, three, and six months for flagged versus
non-flagged stocks.

```{r anomalies}
anom <- anomaly_compare(feats_g)
```

```{r anomaly-fwd-table}
anom$fwd_compare |>
  arrange(scheme, horizon, flagged) |>
  kable(digits = 4,
        caption = "Mean forward alpha for flagged vs. non-flagged stocks, by scheme and horizon")
```

```{r anomaly-fwd-plot}
#| fig-cap: "Forward alpha gap (flagged minus non-flagged) by horizon. A negative bar at +1m would suggest flagged stocks subsequently underperform on a risk-adjusted basis — the mispricing reading."
anom$fwd_compare |>
  group_by(scheme, horizon) |>
  summarise(
    gap = mean_fwd_alpha[flagged] - mean_fwd_alpha[!flagged],
    .groups = "drop"
  ) |>
  ggplot(aes(factor(horizon), gap, fill = scheme)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_hline(yintercept = 0) +
  scale_fill_manual(values = c(GICS = "#1F4E79", FF12 = "#2E75B6", Cluster = "#C00000")) +
  labs(x = "horizon (months ahead)",
       y = "mean forward alpha (flagged − non-flagged)", fill = NULL)
```

```{r anomaly-overlap}
anom$flag_overlap |>
  filter(a != b) |>
  kable(digits = 3,
        caption = "Pairwise Jaccard overlap of flagged (permno, date_eom) keys")
```

# 7. Discussion

The four-dimensional comparison gives a more complete answer than any
single metric:

- **Composition.** Low Rand-index values between K-Means and either of
  the industry schemes confirm that the unsupervised clustering finds
  structure that is not reducible to industry codes. The GICS-FF12 ARI is
  expected to be the highest of the three pairs — both partitions are
  ultimately built around the same notion of what a firm "does."
- **Homogeneity.** Co-movement clusters are constructed to maximise
  within-group return correlation, so they should win on
  `mean_within_corr` by design. The interesting question is whether GICS
  or FF12 produces lower alpha-dispersion within groups — that would
  indicate that industry membership *also* organises unexplained-return
  behaviour, even though the schemes were not built with alpha in mind.
- **Prediction.** The economically meaningful comparison is whether
  per-group models beat the pooled baseline. If they do, the grouping is
  carrying information that conditions the alpha-forecasting relationship
  — different industries respond differently to factor exposures and
  macro conditions. If the pooled model wins, the data is telling us
  there isn't enough sample within any single industry or cluster to
  reliably estimate group-specific dynamics, and one global model is the
  right answer.
- **Anomalies.** A scheme whose Isolation Forest flags subsequently earn
  meaningfully different forward alpha — and where that difference
  persists at 3 and 6 months rather than mean-reverting in a single month
  — has a credible mispricing reading. If all three schemes flag
  similar-quality anomalies, the choice of grouping doesn't matter much
  for outlier detection, and one would prefer the simpler / more
  interpretable scheme (GICS).

Two caveats. First, the dev-mode subset of fifty stocks limits how much
weight the numbers can bear; a full S&P 500 run is necessary before
treating these comparisons as decisive. Second, the alpha computation is
itself dependent on the FF5 specification — alpha that the FF5 model
fails to absorb might still be priced by extensions (momentum, quality,
short-term reversal). Our anomaly definition is "FF5-residual," not
"genuinely unexplained."

# 8. Limitations and next steps

- **Universe size.** Final numbers should come from `CFG$dev_mode = FALSE`
  on the full universe, not the 50-stock sample.
- **Cluster stability.** The annual-refit + Hungarian-matching pipeline
  keeps cluster IDs stable but doesn't guarantee that the underlying
  groupings are stable. A robustness check that varies the refit window
  (say, 36 vs. 60 vs. 84 months) would be useful.
- **Alternative anomaly horizons.** Six months is the longest forward
  horizon we report. Some mispricing-reversion stories play out over a
  year or longer; extending the horizon would test those.
- **Sector-conditional clustering.** A natural follow-up is K-Means
  *within* GICS sectors — does sub-sectoral clustering find structure
  that flat clustering misses?

# Outputs

```{r outputs-table}
list.files(c(PATHS$raw, PATHS$processed),
           pattern = "\\.parquet$", full.names = TRUE) |>
  tibble(file = _) |>
  mutate(size_kb = round(file.size(file) / 1024, 1)) |>
  kable()
```
